---
title: "Rivermile Implementation Workflow"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(leaflet)
library(rivermile)
```

### Get River Mile of a Specific Point

For Klamath flow gage 11530500, get the river mile. 

```{r}
flow_gages <- klamathWaterData::flow_gage |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

klamath_11530500 <- flow_gages |> filter(gage_id == "11530500")

```

There are two ways of calculating the river mile for a single point. 

1) Using `rivermile::get_nearest_river_mile_pt`. This returns a single value of the river mile. 

```{r}
rivermile::get_nearest_river_mile_pt(rivermile::all_klamath_rivers_pts, klamath_11530500)
```

2) Using `rivermile::find_nearest_river_miles`. This returns an updated spatial dataframe with the river mile. 

```{r}
klamath_11530500_with_rm <- rivermile::find_nearest_river_miles(klamath_11530500) |> glimpse()
```

```{r echo=FALSE}
pal_rm <- colorNumeric(
  palette = "viridis",
  domain  = rivermile::klamath_river$river_mile
)

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron, group = "Positron") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Imagery") |> 
  addMarkers(data = klamath_11530500_with_rm, 
             popup = ~paste0("Gage: ", gage_name, "<br>",
                             "River mile: ", river_mile)) |> 
  addCircleMarkers(
    data = rivermile::klamath_river |> st_transform(4326),
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.9,
    fillColor = ~pal_rm(river_mile),
    popup = ~paste0("River mile: ", river_mile),
    group = "River miles"
  ) |> 
  addLegend(
    pal = pal_rm,
    values = rivermile::klamath_river$river_mile,
    title = "River mile",
    opacity = 1
  ) 

```

### Get River Mile Locations from Gage Data

This method calculates river miles for multiple gage locations. When the gage is outside of the package limits, it returns an error message and an NA for `river_mile`

```{r}
temperature <- klamathWaterData::temperature_gage
```

```{r message=FALSE, warning=FALSE}
river_pts <- temperature |> 
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) 

rms <- rivermile::find_nearest_river_miles(river_pts) |> glimpse()
```

```{r echo=FALSE}
pal_rm <- colorNumeric(
  palette = "viridis",
  domain  = rivermile::all_klamath_rivers_pts$river_mile
)

leaflet() |>
  addTiles() |>
  addCircleMarkers(
    data = rivermile::all_klamath_rivers_pts,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.9,
    fillColor = ~pal_rm(river_mile),
    popup = ~paste0("River: ", river, "<br>",
                    "River mile: ", river_mile),
    group = "River miles"
  ) |> 
  addCircleMarkers(data = rms |> filter(!is.na(river_mile)),
                   radius = 8, 
                   fillOpacity = 0.8, 
                   fillColor = "green",
                   color = "darkgreen",
                   popup = ~paste0("River: ", stream, "<br>",
                                   "River mile: ", river_mile, "<br>",
                                   "Gage Name: ", gage_name)) |> 
  addLegend(
    pal = pal_rm,
    values =  rivermile::all_klamath_rivers_pts$river_mile,
    title = "River mile",
    opacity = 1
  ) 


```
